---
title: "Regression Models Course Project"
author: "Joy SN"
date: "January 5, 2019"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# How is MPG affected by Transmission Type (Automatic and Manual)?

## Promblem Summary

Study Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, explored the relationship between a set of variables and miles per gallon (MPG) (outcome). Two questions of importance

"Is an automatic or manual transmission better for MPG"  
"Quantify the MPG difference between automatic and manual transmissions"

### Exploratory Data Analysis
Dataset: Motor Trend Car Road Tests https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html.

```{r echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
# LOAD PACKAGES
library(reshape2)
library(ggplot2)
# DATA SUMMARY
data("mtcars")
#str(mtcars)
data <- mtcars
```

	 mpg	 Miles/(US) gallon  
	 cyl	 Number of cylinders  
	 disp	 Displacement (cu.in.)  
	 hp	 Gross horsepower  
	 drat	 Rear axle ratio  
	 wt	 Weight (1000 lbs)  
	 qsec	 1/4 mile time  
	 vs	 Engine (0 = V-shaped, 1 = straight)  
	 am	 Transmission (0 = automatic, 1 = manual)  
	 gear	 Number of forward gears  
	 carb	 Number of carburetors  

The *figure 1*, we see that correlation between different continuos variables. Per the *Figure 1*, MPG is not so strongly related to drat and qsec

### Statistical Inference

```{r echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
mean_mpg_auto <- mean(data[data$am==0,]$mpg)
mean_mpg_manual <- mean(data[data$am==1,]$mpg)
```

Cars with "manual"" transmission have with mean MPG of `r round(mean_mpg_manual, 1)` miles/gallon
Cars with "automatic" tranmission have mean MPG of  `r round(mean_mpg_auto, 1)` miles/gallon

```{R echo= FALSE}
difference_mpg_am <- t.test(data[data$am==1,]$mpg,data[data$am==0,]$mpg)
# paste0("Lower Range:-", round(difference_mpg_am$conf.int[1],1))
# paste0("Upper Range:-", round(difference_mpg_am$conf.int[2],1))
# paste0("P Value:-", round(difference_mpg_am$p.value,4))
difference_mpg_am
```
Cars with manual transmission Average MPG is 24.39231 miles/gallon compared to  17.14737  MPG in automatic transmission.   
95% CI: `r round(difference_mpg_am$conf.int[1],1)` - `r round(difference_mpg_am$conf.int[2],1)` ,p = `r round(difference_mpg_am$p.value,4)`).

### Model Genration
We start with a model with all the columns as perdictors and based on different P-values we remove each columns

```{r echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
summary(lm(mpg ~ ., data = mtcars))$coefficients
```
Since the P value of cyl is very high, we remove it and build another model minus cyl
```{r echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE, results='hide'}
dat <- mtcars
dat <- dat[, names(dat) != "cyl"]
summary(lm(mpg ~ ., data = dat))$coefficients
```
Then we see that the P value of 'vs' is quite high, so we now build another one without 'VS' and so on we remove 'carb', 'gear', 'drat','disp' and 'hp' because of their high P value
```{r echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE,results='hide'}
dat <- dat[, names(dat) != "vs"]
#summary(lm(mpg ~ ., data = dat))$coefficients
dat <- dat[, names(dat) != "carb"]
#summary(lm(mpg ~ ., data = dat))$coefficients
dat <- dat[, names(dat) != "gear"]
#summary(lm(mpg ~ ., data = dat))$coefficients
dat <- dat[, names(dat) != "drat"]
#summary(lm(mpg ~ ., data = dat))$coefficients
dat <- dat[, names(dat) != "disp"]
#summary(lm(mpg ~ ., data = dat))$coefficients 
dat <- dat[, names(dat) != "hp"]
summary(lm(mpg ~ ., data = dat))$coefficients 
```
As the p-value of all the remaining predictors are smaller than $0.05$, we can stop.
Finally we are leftt with 3 variables - wt, qsec and am  
- am	 Transmission (0 = automatic, 1 = manual)  
- mpg	 Miles/(US) gallon  
- wt	 Weight (1000 lbs)  
- qsec	 1/4 mile time  
```{R echo=FALSE}
dat <- mtcars[, c("mpg", "wt", "qsec", "am")]
summary(lm(mpg ~ . -1, data = dat))$coefficients 
fit <- lm(mpg ~ . -1, data = dat)
```

As the two-sided p-value for the coefficient of `am` is `r summary(fit)$coefficients["am", 4]`, much smaller than 0.05,
we can reject the hypothesis $H_0$.

```{R echo=FALSE}
sumCoef <- summary(fit)$coefficients
intv <- sumCoef["am", 1] + c(-1, 1) * qt(0.975, df = fit$df)*sumCoef["am", 2]
```

With 95% confidence, we estimate that a the change from automatic to manual transmission results in a  `r intv[1]` to `r intv[2]` increase in miles per gallon for the cars.  
Looks like the manual transmission is better than automatic transmission for mpg. Is it causal?


But there is a caveat.
Per *Figure 2* , which is a pair plot among different predictors we see a relation ship between transmission (am) and weight.
Also in *Figure 3*, which is a box plot of, we see there is a strong relation between weight and Transmission

### Final Modelling
We start with Sequential Testing

1.Model 1 - MPG ~ wt (Weight)
```{R echo=FALSE}
fit1 <- lm(mpg ~ wt, data = data)
summary(fit1)$coefficients
```

R Squared :- `r round(summary(fit1)$adj.r.squared,2)`  
SD :- `r round(sd(fit1$residuals),2)`    

2. Model 2 - MPG ~ wt (Weight) + am (Transmission Type)
Next we add column AM (Transmission type)

```{R echo=FALSE}
fit2 <- update(fit1, mpg ~ wt + am)
summary(fit2)$coefficients
```

R Squared :- `r round(summary(fit2)$adj.r.squared,2)`   
SD :- `r round(sd(fit2$residuals),2)`    
We see that the addition of the variable "am"(transmission type) to Model 1 in Model 2 does not change the values much. Thus it is not addition any additional value to explain MPG.

```{R, results = 'hide'}
anv1<-anova(fit1, fit2)
```

Using anova, we get  
RSS Values `r anv1$RSS   `
P Value 0.9879146
```{R echo=FALSE, results='hide'}
anv1$`Pr(>F)`
```
  
  
The effect that "transmission on "MPG"" is mostly due to its relation to weight which was described above (manual cars
are significantly lighter than larger cars)

But as seen in *Figure 4*, there appears to be an interaction between weight and type of transmission to explain MPG. There are different different slopes for weight for each type of transmission, with manual transmission showing a greater decrease in MPG as the weight of the car increases.

3. Model 3 - mpg ~ wt * am 

This model includes the  interaction (weight*transmission)
```{R echo=FALSE}
fit3 <- update(fit1, mpg ~ wt * am)
summary(fit3)$coefficients
anv2<-anova(fit1, fit3)
anv2
```

R Squared :- `r round(summary(fit3)$adj.r.squared,2)`  
RSS Values `r anv2$RSS   `  
In this final model all coefficients are significant  
ANOVA shows that there is some inclusion of wt * am term and it reduces RSS.
  
For "manual" engine cars, for every unit increase in weight(wt) , there is an average drop of MPG = `r round(abs(summary(fit3)$coefficients[2,1])+abs(abs(summary(fit3)$coefficients[4,1])),2)` miles/gallon
For "automatic" cars, this fall in MPG is `r round(abs(summary(fit3)$coefficients[2,1]))` miles/gallon.


*Figure 5* shows the residual and other plots for this final fit.
We note that the residuals show no obvious pattern, so it is reasonable to try to fit a linear model to the data.
Now with all the previous analysis, we can conclude that our linear model is a resonable fit. 


## Conclusion/Executive Summary

MPG of Manual engine cars are on an average less than that of automatic engine tarnsmission type cares. This is also due to the fact that "manual" type cars are lighter than "automatic" engine cars. 
But, for "manual" cars, MPG w.r.t to weight is effected higher (9.08 miles/gallon) than that of "automatic" cars (4 miles/gallon)


## Appendix A
### Fig 1 - Shows the correlation of continuous data
```{r echo=FALSE, fig.width=5, fig.height=3.4}
# Fig 1: - Take the continuous data and plot the correlation diagram
mydata <- mtcars[, c(1,3,4,5,6,7)]
cormat <- round(cor(mydata),2)
melted_cormat <- melt(cormat)
ggplot(data = melted_cormat, aes(x=Var1, y=Var2, fill=value)) + geom_tile()
```

### Fig 2 - Pair Plot among mpg, wt,qsec and am
```{r echo=FALSE, fig.width=5, fig.height=5}
pairs(dat, panel = panel.smooth, main = "mtcars data")
```

### Fig 3 Wights w.r.t MPG for both types of Transmission Type
```{r echo=FALSE, fig.width=3, fig.height=3}
ggplot(mtcars) + 
    geom_boxplot(aes(factor(am), mpg)) + xlab('Weight') +
    scale_x_discrete(labels = c('Manual','Automatic'))
```

### Fig 4 - Shows the Mean wights of both transmission type w.r.t MPG
```{r echo=FALSE, fig.width=5, fig.height=5}
mean_mpg_auto <- mean(data[data$am==0,]$mpg)
mean_mpg_manual <- mean(data[data$am==1,]$mpg)

lmA <- lm(data$mpg[data$am=="0"] ~ data$wt[data$am=="0"])
lmM <- lm(data$mpg[data$am=="1"] ~ data$wt[data$am=="1"])
plot(data$wt,data$mpg,pch=19, xlab='Weight lb/1000', ylab='Miles per gallon')
points(data$wt,data$mpg,pch=19,col=((data$am=="1")*1+1))
lines(data$wt[data$am=="0"],lmA$fitted,col="black",lwd=3)
lines(data$wt[data$am=="1"],lmM$fitted,col="blue",lwd=3)
#points(data$wt,fit3$fitted,col="blue",lwd=2)
abline(h=mean(data$mpg), v=mean(data$wt))
abline(h=mean_mpg_manual, col = "red", lty=2)
abline(h=mean_mpg_auto, lty=2)
text(5,mean(data$mpg[data$am==1])+1,'Mean MPG in manual')
text(5,mean(data$mpg[data$am==0])+1,'Mean MPG in automatic')
```


### Fig 5 - Residual and other variations
```{r echo=FALSE, fig.width=5, fig.height=5}
par(mfrow = c(2, 2))
plot(fit3)
```