# Reproducible Research: Peer Assessment 1

## Data Preperation

### Download and unzip the file

```{r}
my_path <- getwd()
remote_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
downloaded_zip_file = "Activity.zip"
download.file(remote_url, file.path(my_path, downloaded_zip_file))
unzip(zipfile = downloaded_zip_file)
```


### Load Data
```{r}
activity <- read.csv(file.path(my_path, "activity.csv"),header=TRUE)
```

### Data Outlook

```{r}
str(activity)
summary(activity)
head(activity)
```


## What is mean total number of steps taken per day?

### Subset the original data to calculate the total number of steps
```{r}
TotStepsDay <- aggregate(steps~date, activity, sum)
head(TotStepsDay)
```

### Make a histogram of the total number of steps taken each day
```{r}
hist(TotStepsDay$steps,col="red",xlab="Total Steps",main="Total Steps Per Day")
```

### Calculate and report the mean and median total number of steps taken per day
```{r}
mean(TotStepsDay$steps)
median(TotStepsDay$steps)
```

## What is the average daily activity pattern?

### Calculate the mean(average) per interval
```{r}
AvgStepsInt <- aggregate(steps~interval, activity, mean)
head(AvgStepsInt)
```

### Make a time series plot (type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
with(AvgStepsInt, plot(interval, steps, type="l", xlab = "Intreval", ylab= "Number of Steps", main="Average number of steps taken in 5-min interval", col = "red")) 
```

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max <- max(AvgStepsInt$steps)
max
MaxAvgStepsInt <- subset(AvgStepsInt, steps == max)
MaxAvgStepsInt$interval
```

## Inputing missing values.
Note that there are a number of days/intervals where there are missing values (NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
colSums(is.na(activity))
```

2. Devise a strategy for filling in all of the missing values in the dataset. We are filling the mean for that day, or the mean for that 5-minute interval.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
activityNew <- activity
activityNew$steps[is.na(activityNew$steps)] <- AvgStepsInt$steps
head(activityNew)
```

4. Make a histogram of the total number of steps taken each day 
```{r}
TotStepsDayNew <- aggregate(steps~date, activityNew, sum)
hist(TotStepsDayNew$steps,col="red",xlab="Total Steps",main="Total Steps Per Day")
```

5. Calculate and report the mean and median total number of steps taken per day. 
```{r}
mean(TotStepsDayNew$steps)
median(TotStepsDayNew$steps)
```

6. Do these values differ from the estimates from the first part of the assignment?  
  What is the impact of imputing missing data on the estimates of the total daily number of steps?  
  Not much inpact seen in mean/median of the data. Mean remains same, median varied very little
```{r}
mean(TotStepsDayNew$steps) - mean(TotStepsDay$steps)
median(TotStepsDayNew$steps) - median(TotStepsDay$steps)
```

## Are there differences in activity patterns between weekdays and weekends?
Use the dataset with the filled-in missing values for this part.

1. Install the package  
  install.packages("timeDate")

2. Load
```{r}
library(timeDate)
```

3. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
activityNew$Weekday <- factor(isWeekday(activityNew$date))
str(activityNew)
head(activityNew)
```

### Make a panel plot containing a time series plot (i.e. type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

1. Subset and calculate the average steps for weekday and weekend
```{r}
activityWeekday <- subset(activityNew, activityNew$Weekday == "TRUE")
AvgWeekdayStepsInt <- aggregate(steps~interval, activityWeekday, mean)
head(AvgWeekdayStepsInt)
```

2. Subset and calculate the average steps for weekend
```{r}
activityWeekend <- subset(activityNew, activityNew$Weekday == "FALSE")
AvgWeekendStepsInt <- aggregate(steps~interval, activityWeekend, mean)
head(AvgWeekendStepsInt)
```

3. Ploting the Graph

```{r}
par(mfrow = c(2,1))

layout(matrix(c(1,1,2,2), 2, 2, byrow = TRUE))
# Weekday
plot(AvgWeekdayStepsInt$interval, AvgWeekdayStepsInt$steps, xlab = "Interval", ylab = "Number of Steps", main ="Weekday", col ="blue", type="l") 
# Weekend
plot(AvgWeekendStepsInt$interval, AvgWeekendStepsInt$steps, xlab = "Interval", ylab = "Number of Steps", main ="Weekend", col ="red", type="l")
```